MG7: Configurable and scalable 16S data analysis

### 2016-07-26 docs/mg7-paper/out/paper.tex
fix named links in paper

\begin{itemize}
\tightlist
\item
  Static reproducible specification of dependencies and behavior of the
  different components using
  \emph{\href{https://github.com/ohnosequences/statika}{Statika}} and
  \emph{\href{https://github.com/ohnosequences/datasets}{Datasets}}
\item
  Parallelization and distributed analysis based on AWS, with on-demand
  infrastructure as the basic paradigm
\item
  Definition of complex workflows using
  \emph{\href{https://github.com/ohnosequences/loquat}{Loquat}}, a
  composable system for scaling/parallelizing stateless computations
  especially designed for AWS
\item
  A new approach to data analysis specification, management and
  specification based on working with it in exactly the same way as for
  a software project, together with the extensive use of compile-time
  structures and checks
\item
  Modeling of the taxonomy tree using the new paradigm of graph
  databases (Bio4j, \citep{pareja2015bio4j}). It facilitates the
  taxonomic assignment tasks and the calculation of the taxa abundance
  values considering the hierarchic structure of taxonomy tree
\item
  Exhaustive per-read taxonomic assignment using two complementary
  assignment algorithms: Lowest Common Ancestor (LCA) and Best BLAST Hit
\item
  Using a new 16S database of reference sequences (16S-DB7) with a
  flexible and sustainable system of updating and project-driven
  customization
\end{itemize}

@@

The second step is to search for similar 16S sequences in our 16S-DB7
database. The taxonomic assignment for each read is based on BLASTN of
each read against the 16S database.Assignment based on direct similarity
of each read one by one compared against a sufficiently wide database is
considered in different reviews of metagenomics analysis methodologies
\citep{segata2013computational} \citep{morgan2012chapter} as a very
exhaustive method for assignment. Some methods of assignment compare the
sequences only against the 16S genes from available complete bacterial
genomes or avoid computational cost clustering or binning the sequences
first, and then doing the assignments only for the representative
sequence of each cluster. MG7 carries out an exhaustive comparison of
all the reads under analysis and it does not applies any binning
strategy. Every read is specifically compared with all the sequences of
the 16S database.

@@

All the reads are assigned under two different algorithms of assignment:
i. Lowest Common Ancestor based taxonomic assignment (LCA) and ii. Best
BLAST Hit based taxonomic assignment (BBH). Figure 2 displays
schematically the LCA algorithm.

\begin{figure}[htbp]
\centering
\includegraphics{./Figure-2.jpg}
\caption{\textbf{Lowest Common Ancestor algorithm for taxonomic
assignment.} \textbf{TODO UPDATE THIS FIGURE!} The Left panel displays
an example of the application of LCA algorithm in a \emph{sensu stricto}
mode. A, B, C and D represent taxonomy tree nodes with assigned reads.
Right panel displays the \emph{in line} mode of assignment which is an
exception for the \emph{sensu stricto} mode of application of LCA
algorithm. The \emph{in line} mode is used when all the nodes are
located in a line without bifurcations. In that case the taxon assigned
is the most specific (the most distant from the root).}
\end{figure}

All the assignment procedures work by default with input the set of
BLAST hits which have maximum bitscore among those covering the whole
query sequence; we will call this the set of valid hits of a query. Note
that the criteria which determines this set of valid hits is fully
configurable, and should be adjusted to conform to its meaning in each
particular dataset: those hits which we should take into account as
valid for assignment. This is of course experiment-dependent, and the
criteria for valid hits should account for read quality or
technology-specific error profiles.

\paragraph{LCA Assignment}\label{lca-assignment}

The LCA assignment simply computes the lowest common ancestor of the set
of valid hits. Other metagenomics analysis approaches
\citep{huson2012microbial} have adopted LCA-based assignment algorithms,
as it provides the most precise assignment for a set of taxa, with no
further information available\footnote{the lowest common ancestor of a
  set of nodes in a tree is just their product in the tree viewed as a
  poset, with the ``parent of'' partial order}; as our reference
database is already heavily preprocessed and checked for internal
consistency, there is no need for altering the lowest common ancestor in
any way, as is customarily done in other cases. MG7 operates under the
assumption that hits which fulfill the user-defined criteria for being
valid hits should be treated on an equal footing, and in that case
taking their LCA is the coherent assignment choice with minimum
information loss.

>>>>

\begin{itemize}
\tightlist
\item
  Static reproducible specification of dependencies and behavior of the
  different components using
  \emph{\href{https://github.com/ohnosequences/statika}{Statika}} and
  \emph{\href{https://github.com/ohnosequences/datasets}{Datasets}}
\item
  Parallelization and distributed analysis based on AWS, with on-demand
  infrastructure as the basic paradigm
\item
  Definition of complex workflows using
  \emph{\href{https://github.com/ohnosequences/loquat}{Loquat}}, a
  composable system for scaling/parallelizing stateless computations
  especially designed for AWS
\item
  A new approach to data analysis specification, management and
  specification based on working with it in exactly the same way as for
  a software project, together with the extensive use of compile-time
  structures and checks
\item
  Modeling of the taxonomy tree using the new paradigm of graph
  databases (Bio4j, \citep{pareja2015bio4j}). It facilitates the
  taxonomic assignment tasks and the calculation of the taxa abundance
  values considering the hierarchic structure of taxonomy tree
\item
  Exhaustive per-read taxonomic assignment using two complementary
  assignment algorithms Lowest Common Ancestor and Best BLAST Hit
\item
  Using a new 16S database of reference sequences (16S-DB7) with a
  flexible and sustainable system of updating and project-driven
  customization
\end{itemize}

@@

The second step is to search for similar 16S sequences in our 16S-DB7
database. The taxonomic assignment for each read is based on BLASTN of
each read against the 16S database.Assignment based on direct similarity
of each read one by one compared against a sufficiently wide database is
considered in different reviews of metagenomics analysis methodologies
\citep{segata2013computational} \citep{morgan2012chapter} as a very
exhaustive method for assignment. Some methods of assignment compare the
sequences only against the 16S genes from available complete bacterial
genomes or avoid computational cost clustering or binning the sequences
first, and then doing the assignments only for the representative
sequence of each cluster. MG7 carries out an exhaustive comparison of
all the reads under analysis and it does not applies any binning
strategy. Every read is specifically compared with all the sequences of
the 16S database. We select the best BLAST hits (10 hits by default)
obtained for each read to do the taxonomic assignment.

@@

All the reads are assigned under two different algorithms of assignment:
i. Lowest Common Ancestor based taxonomic assignment (LCA) and ii. Best
BLAST Hit based taxonomic assignment (BBH). Figure 2 displays
schematically the LCA algorithm applied sensu stricto (left panel) and
the called `in line' exception (right panel) designed in order to gain
specificity in the assignments in the cases in which the topology of the
taxonomical nodes corresponding to the BLAST hits support sufficiently
the assignment to the most specific taxon.

\begin{figure}[htbp]
\centering
\includegraphics{./Figure-2.jpg}
\caption{\textbf{Lowest Common Ancestor algorithm for taxonomic
assignment.} The Left panel displays an example of the application of
LCA algorithm in a \emph{sensu stricto} mode. A, B, C and D represent
taxonomy tree nodes with assigned reads. Right panel displays the
\emph{in line} mode of assignment which is an exception for the
\emph{sensu stricto} mode of application of LCA algorithm. The \emph{in
line} mode is used when all the nodes are located in a line without
bifurcations. In that case the taxon assigned is the most specific (the
most distant from the root).}
\end{figure}

\paragraph{Lowest Common Ancestor based Taxonomic
Assignment}\label{lowest-common-ancestor-based-taxonomic-assignment}

For each read, first, we select the BEST BLAST HITs (by default 10 Hits)
over a threshold of similarity. To evaluate similarity for this first
filtering of hits we use the Expect value (by default
\(evalue \leq e^{-15}\)) that describes the number of hits one can
``expect'' to see by chance when searching a database of a particular
size. In a second filtering step we filtering those hits that are not
sufficiently good comparing them with the best one. We select the best
HSP (High Similarity Pair) per reference sequence and then choose the
best HSP (that with lowest E-value) between all the selected ones. The
bitscore of this best HSP (called S) is used as reference to filter the
rest of HSPs. All the HSPs with bitscore below the product \(p S\) are
filtered. p is a coefficient fixed by the user to define the bitscore
required, e.g.~if \(p=0.9\) and \(S=700\) the required bitscore
threshold would be \(630\). Once we have the definitive HSPs selected,
we obtain their corresponding taxonomic nodes using the taxonomic
assignments that NCBI provides for all the nt database sequences. Now we
have to analyze the topological distribution of these nodes in the
taxonomy tree: i. If all the nodes forms a line in the taxonomy tree
(are located in a not branched lineage to the tree root) we should
choose the most specific taxID as the final assignment for that read. We
call to this kind of assignment the `in line' exception (see Figure 2
right panel). ii. If not, we should search for the \emph{sensu stricto}
Lowest Common Ancestor (LCA) of all the selected taxonomic nodes (See
Figure 2 left panel). In this approach we decided to use the bitscore
for evaluating the similarity because it is a value that increases when
similarity is higher and depends a lot on the length of the HSP. Some
reads could not find sequences with enough similarity in the database
and then they would be classified as reads with no hits. Advanced
metagenomics analysis approaches \citep{huson2012microbial} have adopted
LCA-based assignment algorithms because it provides fine and trusted
taxonomical assignment.
